---
title: "Advanced Survival Analysis with jSurvival"
author: "ClinicoPath Development Team"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# Overview

This vignette covers advanced survival analysis techniques and methodological considerations when using jSurvival. It addresses complex scenarios that researchers commonly encounter in clinical and epidemiological studies.

# Advanced Cox Regression Topics

## Checking Proportional Hazards Assumption

The Cox proportional hazards model assumes that hazard ratios remain constant over time. Violation of this assumption can lead to misleading results.

### Visual Assessment

```{r ph-visual}
# Use survival analysis with extended follow-up time
ph_check <- survival(
  data = mydata,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = treatment,
  ph_cox = TRUE,
  endplot = 120,  # Extended follow-up for visual inspection
  sc = TRUE       # Schoenfeld residuals (if available)
)
```

**Visual Signs of PH Violation**:
- Crossing survival curves
- Changing hazard ratios over time
- Non-random patterns in Schoenfeld residuals

### Solutions for PH Violations

#### Option 1: Stratified Cox Model

```{r stratified-cox}
# When PH assumption is violated for a categorical variable
# Use stratification approach in multivariable analysis
stratified_result <- multisurvival(
  data = mydata,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = c(treatment, age, sex),
  # Note: Stratification would need to be implemented
  # This is conceptual - actual implementation may vary
)
```

#### Option 2: Time-Dependent Effects

For variables with time-varying effects, consider:
- Landmark analyses at multiple time points
- Piecewise exponential models
- Flexible parametric models

## Model Selection and Variable Selection

### Forward/Backward Selection

```{r variable-selection}
# Start with univariate screening
candidate_vars <- c("age", "sex", "stage", "grade", "biomarker")
univariate_results <- list()

for(var in candidate_vars) {
  univariate_results[[var]] <- survival(
    data = mydata,
    elapsedtime = time_months,
    outcome = death,
    outcomeLevel = "1",
    explanatory = !!sym(var),
    ph_cox = TRUE
  )
}

# Variables with p < 0.20 in univariate analysis
# proceed to multivariable model
significant_vars <- c("age", "stage", "biomarker")  # Based on results

final_model <- multisurvival(
  data = mydata,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = significant_vars
)
```

### Model Validation

```{r model-validation}
# Internal validation using bootstrap or cross-validation
# External validation in independent dataset

validation_result <- multisurvival(
  data = validation_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = final_model_variables
)
```

## Handling Missing Data

### Complete Case Analysis

```{r complete-case}
# Default approach - uses only complete cases
complete_result <- survival(
  data = mydata,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = treatment
)
```

### Multiple Imputation Approach

```{r multiple-imputation}
# Conceptual approach for multiple imputation
# 1. Create multiple imputed datasets
# 2. Analyze each dataset separately
# 3. Pool results using Rubin's rules

# Example workflow (implementation details may vary):
imputed_results <- list()
for(i in 1:5) {  # 5 imputed datasets
  imputed_results[[i]] <- survival(
    data = imputed_data[[i]],
    elapsedtime = time_months,
    outcome = death,
    outcomeLevel = "1",
    explanatory = treatment
  )
}

# Pool results (conceptual)
pooled_hr <- mean(sapply(imputed_results, function(x) x$hazard_ratio))
pooled_se <- sqrt(mean(sapply(imputed_results, function(x) x$se^2)) + 
                  (1 + 1/5) * var(sapply(imputed_results, function(x) x$hazard_ratio)))
```

# Cut-point Optimization

## Methodological Considerations

### Multiple Cut-point Testing

There is **no p-value adjustment option** on `survivalcont`. `padjustmethod` does not exist and
never has; passing it raises `unused argument`. What the analysis does offer is control over how
the cut-offs themselves are derived, via `multiple_cutoffs`, `num_cutoffs` and `cutoff_method`:

```{r multiple-cutpoints}
cutpoint_analysis <- survivalcont(
  data = biomarker_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  contexpl = biomarker_value,
  findcut = TRUE,           # single cut-off, maximally selected rank statistic
  multiple_cutoffs = TRUE,  # or several candidate cut-offs
  num_cutoffs = "two",      # "two" (3 groups), "three" (4), "four" (5)
  cutoff_method = "quantile" # "quantile", "recursive", "tree", "minpval"
)
```

### Cross-Validation of Cut-points

```{r cutpoint-validation}
# Split data into training and validation sets
set.seed(123)
train_idx <- sample(nrow(biomarker_data), nrow(biomarker_data) * 0.7)
train_data <- biomarker_data[train_idx, ]
test_data <- biomarker_data[-train_idx, ]

# Find cut-point in training data
train_cutpoint <- survivalcont(
  data = train_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  contexpl = biomarker_value,
  findcut = TRUE
)

# Apply cut-point to test data
test_data$biomarker_high <- ifelse(test_data$biomarker_value >= optimal_cutpoint, 
                                   "High", "Low")

validation_result <- survival(
  data = test_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = biomarker_high
)
```

## Alternative Approaches to Dichotomization

### Tertiles or Quartiles

```{r tertiles}
# Instead of single cut-point, use tertiles
biomarker_data$biomarker_tertile <- cut(biomarker_data$biomarker_value,
                                        breaks = quantile(biomarker_data$biomarker_value,
                                                         c(0, 1/3, 2/3, 1), na.rm = TRUE),
                                        labels = c("Low", "Medium", "High"),
                                        include.lowest = TRUE)

tertile_analysis <- survival(
  data = biomarker_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = biomarker_tertile,
  analysistype = "overall"
)
```

### Spline Analysis

Restricted cubic splines are built in — `survival()` gained `rcs_analysis` for exactly this, so
there is no need to leave jsurvival to avoid dichotomising a continuous predictor:

```{r spline-concept}
spline_analysis <- survival(
  data = histopathology,
  elapsedtime = "OverallTime", outcome = "Outcome", outcomeLevel = "1",
  explanatory = "Group",
  dod = "", dooc = "", awd = "", awod = "",
  rcs_analysis = TRUE,
  rcs_variable = "Age",   # the continuous predictor to model flexibly
  rcs_knots    = 4        # 3 knots for a modest cohort, 5 when data allow
)
```

This tests whether the log-hazard really is linear in the predictor. If it is not, a
dichotomisation chosen from the same data will look better than it is. See
[Features Added Since the Earlier Releases](#features-added-since-the-earlier-releases) for the
calibration and internal-validation options that pair with it.

# Time-Dependent Covariates

## Landmark Analysis Implementation

### Multiple Landmark Points

```{r multiple-landmarks}
# Analyze survival at different landmark times
landmark_times <- c(6, 12, 24)  # months
landmark_results <- list()

for(t in landmark_times) {
  # Create landmark dataset
  landmark_data <- subset(original_data, 
                         time_months > t | (time_months <= t & death == 0))
  
  # Adjust survival times
  landmark_data$time_from_landmark <- pmax(0, landmark_data$time_months - t)
  
  # Analyze
  landmark_results[[paste0("Month_", t)]] <- survival(
    data = landmark_data,
    elapsedtime = time_from_landmark,
    outcome = death,
    outcomeLevel = "1",
    explanatory = response_status,
    uselandmark = TRUE,
    landmark = t
  )
}
```

### Dynamic Prediction

```{r dynamic-prediction}
# Conditional survival probabilities
# Probability of surviving additional 2 years given survival to 1 year

# Patients alive at 12 months
alive_12m <- subset(mydata, time_months > 12 | (time_months <= 12 & death == 0))
alive_12m$time_conditional <- pmax(0, alive_12m$time_months - 12)

conditional_survival <- singlearm(
  data = alive_12m,
  elapsedtime = time_conditional,
  outcome = death,
  outcomeLevel = "1",
  cutp = "24",  # Additional 2 years
  timetypeoutput = "months"
)
```

# Competing Risks Analysis

## Cause-Specific Hazards

```{r cause-specific}
# Separate analyses for different causes of death
# Cancer-specific mortality
cancer_death <- survival(
  data = competing_data,
  elapsedtime = time_months,
  outcome = death_cancer,  # 1 = cancer death, 0 = alive or other death
  outcomeLevel = "1",
  explanatory = treatment,
  ph_cox = TRUE
)

# Other-cause mortality
other_death <- survival(
  data = competing_data,
  elapsedtime = time_months,
  outcome = death_other,   # 1 = other death, 0 = alive or cancer death
  outcomeLevel = "1",
  explanatory = treatment,
  ph_cox = TRUE
)
```

## Subdistribution Hazards (Fine-Gray Model)

```{r fine-gray-concept}
# Fine-Gray model for cumulative incidence
# This approach treats competing events as censoring at the time they occur
# but keeps subjects in the risk set

# Implementation would require specialized competing risks functions
# Currently beyond basic jSurvival functionality
```

# Sample Size and Power Calculations

## Post-hoc Power Analysis

```{r power-analysis}
# Given observed data, calculate achieved power
observed_events <- 120
observed_hr <- 0.75
alpha <- 0.05

# Use standard formulas or specialized software for power calculation
# Example calculation (conceptual):
log_hr <- log(observed_hr)
se_log_hr <- 1.96 / abs(qnorm(alpha/2))  # Approximate from CI
power <- pnorm(abs(log_hr)/se_log_hr - qnorm(1-alpha/2))
```

## Required Sample Size

```{r sample-size}
# For planning future studies
target_hr <- 0.70        # Clinically meaningful difference
alpha <- 0.05            # Type I error
power <- 0.80            # Desired power
accrual_time <- 24       # months
followup_time <- 36      # months
median_survival <- 30    # months in control group

# Use specialized software or formulas for calculation
# Required events ≈ 4 * (Z_α/2 + Z_β)² / (log(HR))²
```

# Meta-Analysis of Survival Data

## Individual Patient Data Meta-Analysis

```{r ipd-meta}
# Combine multiple datasets
combined_data <- rbind(
  transform(study1_data, study = "Study1"),
  transform(study2_data, study = "Study2"),
  transform(study3_data, study = "Study3")
)

# Stratified analysis by study
meta_result <- multisurvival(
  data = combined_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = c(treatment, age, sex, study)
  # Consider stratification by study
)
```

## Fixed Effects vs. Random Effects

```{r meta-approaches}
# Fixed effects: assumes same treatment effect across studies
fixed_effects <- survival(
  data = combined_data,
  elapsedtime = time_months,
  outcome = death,
  outcomeLevel = "1",
  explanatory = treatment
)

# Random effects: allows heterogeneity between studies
# Would require specialized meta-analysis functions
```

# Quality Control and Validation

## Data Quality Checks

```{r quality-checks}
# Check for data inconsistencies
data_checks <- list(
  # Negative survival times
  negative_times = sum(mydata$time_months < 0, na.rm = TRUE),
  
  # Events after last follow-up
  impossible_events = sum(mydata$death == 1 & mydata$time_months > max_followup),
  
  # Missing key variables
  missing_time = sum(is.na(mydata$time_months)),
  missing_event = sum(is.na(mydata$death)),
  
  # Extreme values
  extreme_times = sum(mydata$time_months > 200, na.rm = TRUE)  # >16 years
)

print(data_checks)
```

## Model Diagnostics

```{r model-diagnostics}
# Concordance index (C-index)
# Measures discriminative ability of the model
# Values >0.7 generally considered good

# Calibration assessment
# Agreement between predicted and observed survival

# Model comparison using AIC/BIC
# Lower values indicate better fit
```

# Reporting and Interpretation

## Effect Size Interpretation

```{r effect-interpretation}
# Hazard Ratio Interpretation:
# HR = 0.50: 50% reduction in hazard (strong effect)
# HR = 0.75: 25% reduction in hazard (moderate effect)
# HR = 0.90: 10% reduction in hazard (small effect)
# HR = 1.00: No effect
# HR = 1.25: 25% increase in hazard (moderate harm)
# HR = 2.00: 100% increase in hazard (strong harm)

# Number Needed to Treat (NNT) calculation
survival_control <- 0.60  # 5-year survival in control
survival_treatment <- 0.70  # 5-year survival in treatment
absolute_benefit <- survival_treatment - survival_control
nnt <- 1 / absolute_benefit  # Number needed to treat
```

## Confidence Interval Interpretation

```{r ci-interpretation}
# HR = 0.75 (95% CI: 0.60-0.95)
# Interpretation:
# - Point estimate suggests 25% reduction in hazard
# - We can be 95% confident the true HR is between 0.60 and 0.95
# - Since CI excludes 1.0, result is statistically significant
# - Minimum plausible benefit is 5% (HR=0.95)
# - Maximum plausible benefit is 40% (HR=0.60)
```

# Conclusion

Advanced survival analysis requires careful consideration of:

1. **Model assumptions** and their validation
2. **Missing data patterns** and appropriate handling
3. **Multiple comparisons** and adjustment strategies
4. **Clinical relevance** beyond statistical significance
5. **Robust validation** in independent datasets

Key principles for advanced analysis:

- **Pre-specify analysis plans** to avoid data dredging
- **Validate findings** in independent cohorts when possible
- **Consider clinical context** in statistical decisions
- **Report limitations** and assumptions clearly
- **Collaborate with statisticians** for complex analyses


# Features Added Since the Earlier Releases

These options are all live in `survival()` but were not covered anywhere in this guide until now.
Every call below was run against the bundled `histopathology` data before being written down.

## Weighted Log-Rank Tests

The standard log-rank test weights every event equally, which makes it least sensitive exactly
where survival curves often separate first. `weightedLogRank = TRUE` adds the Fleming-Harrington
family alongside it, so you can see whether a conclusion depends on the weighting.

```{r weighted-logrank}
res <- survival(
  data = histopathology,
  elapsedtime = "OverallTime", outcome = "Outcome", outcomeLevel = "1",
  explanatory = "Group",
  dod = "", dooc = "", awd = "", awod = "",
  weightedLogRank  = TRUE,
  survivalTestType = "fh_rho1"   # "logrank", "fh_rho0_5", "fh_rho1"
)
res$weightedLogRankTable$asDF
#>                                   test rho    chisq df   pvalue
#> 1                  Log-Rank (standard) 0.0 0.220312  1 0.638802
#> 2         Fleming-Harrington (rho=0.5) 0.5 0.128574  1 0.719915
#> 3 Peto-Peto (Fleming-Harrington rho=1) 1.0 0.066619  1 0.796325
```

All three tests are reported whichever `survivalTestType` you pick, each with the weighting it
applies: `rho = 0` weights all time points equally, `rho = 0.5` gives moderate emphasis to early
differences, and `rho = 1` weights by S(t) and emphasises early events. Choose the test **before**
looking at the results — picking the smallest p-value from three tests is not a test.

## Age-Adjusted Analysis

Age is the dominant prognostic variable in most clinical cohorts, and an unadjusted group
comparison can be reporting age rather than treatment. Seven options cover this:

```{r age-adjusted}
res_age <- survival(
  data = histopathology,
  elapsedtime = "OverallTime", outcome = "Outcome", outcomeLevel = "1",
  explanatory = "Group",
  dod = "", dooc = "", awd = "", awod = "",
  age_adjustment = TRUE,
  age_variable   = "Age"
)
res_age$ageAdjustedCoxTable$asDF
#>   variable            levels   n  hr_unadjusted hr_age_adjusted
#> 1    group         Treatment 245 0.94 (0.69-1.27) 0.94 (0.69-1.28)
#> 2      Age per unit (p=0.11)  NA                - 0.99 (0.98-1.00)
```

The table puts the unadjusted and age-adjusted hazard ratios side by side, which is the
comparison that actually answers "is this confounded by age?". The remaining options address
different questions and can be combined with the above:

| Option | Question it answers |
|---|---|
| `age_interaction = TRUE` | Does the treatment effect differ by age? |
| `age_stratified_cox = TRUE` | Adjust for age without assuming proportional hazards across age |
| `age_group_cutpoints` | Age bands for stratification (default `"50, 65, 75"`) |
| `age_time_scale = TRUE` | Use age, not time-on-study, as the time scale |
| `age_standardization = TRUE` | Standardised mortality ratio; `age_standardization_method` is `"indirect"` or `"direct"` |
| `age_stratified_km = TRUE` | Kaplan-Meier curves per age band |

## Parametric Survival Models

Kaplan-Meier and Cox are non-parametric and semi-parametric respectively; neither extrapolates
beyond the observed follow-up. When you need to, `use_parametric = TRUE` fits a parametric model.

```{r parametric}
res_par <- survival(
  data = histopathology,
  elapsedtime = "OverallTime", outcome = "Outcome", outcomeLevel = "1",
  explanatory = "Group",
  dod = "", dooc = "", awd = "", awod = "",
  use_parametric          = TRUE,
  parametric_distribution = "weibull",   # exp, weibull, lnorm, llogis, gamma,
                                         # gengamma, gompertz, survspline
  parametric_covariates   = TRUE,
  compare_distributions   = TRUE         # fit several and compare by information criterion
)
```

With `parametric_distribution = "survspline"`, `spline_knots` (default 3) and `spline_scale`
(`"hazard"`, `"odds"` or `"normal"`) control the flexible-parametric fit. Extrapolation beyond the
follow-up window is a modelling assumption, not a measurement — say so when you report it.

## Calibration, Non-Linearity and Internal Validation

```{r validation-options}
res_val <- survival(
  data = histopathology,
  elapsedtime = "OverallTime", outcome = "Outcome", outcomeLevel = "1",
  explanatory = "Group",
  dod = "", dooc = "", awd = "", awod = "",
  calibration_curves    = TRUE,   # predicted vs observed survival
  calibration_timepoint = 12,     # in the analysis time unit
  calibration_ngroups   = 5,      # risk groups on the calibration plot
  rcs_analysis          = TRUE,   # restricted cubic splines: is the effect linear?
  rcs_variable          = "Age",
  rcs_knots             = 4,
  bootstrapValN         = 200     # bootstrap resamples for internal validation
)
```

`rcs_analysis` answers a question that is easy to skip: a continuous predictor entered linearly
assumes the log-hazard is linear in it, and that is frequently false for age and for biomarkers.
Bootstrap validation here is *internal* — it corrects for optimism, it does not substitute for an
independent cohort.

## Model Performance in `multisurvival`

Four `multisurvival()` options address the question a Cox table cannot answer on its own — how well
does this model actually predict? All four ran against the bundled `histopathology` data.

```{r multisurvival-performance}
res_ms <- multisurvival(
  data = histopathology,
  elapsedtime = "OverallTime", outcome = "Outcome", outcomeLevel = "1",
  explanatory = c("Group", "LVI"), contexpl = "Age",
  dod = "", dooc = "", awd = "", awod = "",
  ci_optimism      = TRUE,   # optimism-corrected C-index by bootstrap
  ci_optimism_boot = 200,    # resamples (default 150)
  compare_models   = TRUE,   # covariate contribution by single-term deletion
  show_survmetrics = TRUE,   # Brier score and time-dependent AUC
  survmetrics_timepoints = "12, 36, 60",
  survmetrics_show_plots = TRUE
)
```

| Option | What it gives you |
|---|---|
| `ci_optimism` | The apparent C-index of a model fitted and evaluated on the same rows is optimistic. This bootstraps the whole fitting procedure to estimate that optimism and subtracts it. Report the corrected figure. |
| `compare_models` | Drops each term in turn and reports the change in fit, which answers "what is this covariate contributing?" more honestly than reading its p-value. |
| `show_survmetrics` | Brier score (calibration and discrimination together, lower is better) and time-dependent AUC at the timepoints you name in `survmetrics_timepoints`. |
| `ac_summary` | With `ac = TRUE` and `adjexplanatory` set, adds adjusted probability summary tables to the adjusted survival curves. |

Optimism correction is still *internal* validation. It tells you how much this modelling procedure
overfits these data; it cannot tell you how the model behaves in another hospital's cohort.

## Reporting Aids

```{r reporting-aids}
res_report <- survival(
  data = histopathology,
  elapsedtime = "OverallTime", outcome = "Outcome", outcomeLevel = "1",
  explanatory = "Group",
  dod = "", dooc = "", awd = "", awod = "",
  adjusted_curves  = TRUE,   # covariate-adjusted survival curves
  remark_checklist = TRUE,   # REMARK reporting checklist for prognostic marker studies
  showSummaries    = TRUE    # natural-language summaries alongside the tables
)
```

`remark_checklist` is worth running on any prognostic-marker analysis you intend to publish; it
lists the REMARK items and is quicker than reconstructing them from the paper afterwards.


For complex analyses beyond the scope of basic jSurvival functions, consider:
- Specialized R packages (survival, survminer, rms, cmprsk)
- Statistical software with advanced survival capabilities
- Consultation with biostatisticians

Additional resources:
- [Survival Analysis Handbook](https://www.emilyzabor.com/tutorials/survival_analysis_in_r_tutorial.html)
- [Clinical Prediction Models](https://www.clinicalpredictionmodels.org/)
- [STROBE Guidelines](https://www.strobe-statement.org/) for observational studies
