---
title: "Interaction terms in Multivariable Survival Analysis"
subtitle: "Testing effect modification: does treatment benefit differ by biomarker?"
author: "ClinicoPath"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interaction terms in Multivariable Survival Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  eval = FALSE
)
```

## When to use this

A randomized trial rarely stops at "does the treatment work?" The next
question is *"who does it work for?"* - does the treatment effect differ
across a biomarker, a stage, or a histologic subtype? This is **effect
modification**, and the standard way to test it in a Cox model is a
**Treatment × Biomarker interaction term**:
`Surv(time, event) ~ Treatment * Biomarker`.

`multisurvival` (Multivariable Survival Analysis) now supports building these
interaction terms directly, without leaving jamovi.

**Interaction is not the same as stratification.** `multisurvival` already
supports `strata()` for stratified Cox models: stratifying by a variable
gives each stratum its own baseline hazard, but it *forces the covariate
effects to be identical across strata* - it cannot tell you whether a
treatment effect differs by subgroup. An interaction term does the opposite:
it estimates a coefficient for the crossed term and directly tests whether
the effect is modified. If your question is "does the biomarker predict who
benefits from treatment?", you need an interaction term, not (only)
stratification.

## Building an interaction term in the GUI

1. Assign variables to **Explanatory Variables** and/or **Continuous
   Explanatory Variable** as usual - these become the main effects in the
   Cox model.
2. Open the **Interaction Terms** section (collapsed by default). It lists
   every variable already chosen as a main effect as an available predictor.
3. Select two (or more) of those variables and use the transfer button to
   cross them into a term, exactly like jamovi's regression/ANOVA Model
   Terms builders.

Because only variables already selected as main effects can be crossed, the
builder enforces the **marginality (hierarchy) principle** by construction - 
you cannot add `Treatment:Biomarker` without both `Treatment` and
`Biomarker` already present as main effects.

> **Convention:** for a two-way term `A × B`, **A (listed first) is the focal
> effect** and **B (listed second) is the moderator**. The focal effect is the
> variable whose hazard ratio you want to see change across subgroups (e.g.
> Treatment); the moderator defines the subgroups (e.g. Biomarker status). To
> get the effect in the other direction, build `B × A` instead.

## What you get

**1. The standard outputs, extended.** Interaction terms appear as
additional rows in the main hazard-ratio table and in the forest plot,
alongside the main effects - no separate step needed.

**2. Interaction (Effect-Modification) Test.** A dedicated table with one
row per interaction term: hazard ratio, 95% CI, and the interaction p-value.
A significant p indicates the focal effect genuinely differs across levels
of the moderator (effect modification) - this is the formal test
biomarker-by-treatment analyses report.

**3. Within-Subgroup Hazard Ratios.** For each two-way interaction whose
moderator is **categorical**, `multisurvival` releveles the data to each
moderator level in turn and refits the *same full* Cox model (all
covariates and interactions, unchanged), then reports the focal variable's
hazard ratio within that subgroup. This gives the fully-adjusted "HR for
Treatment in Biomarker-positive patients" / "HR for Treatment in
Biomarker-negative patients" rows clinicians actually want to read, computed
from one coherent model rather than separate subgroup-only fits.

## Limitations

- **Within-subgroup HRs need a categorical moderator.** If the moderator is
  continuous, the interaction coefficient in the effect-modification test
  table is still reported, but per-subgroup HRs are not computed (there is
  no discrete "subgroup" to relevel to) - a note explains this in the
  output.
- **Within-subgroup HRs are computed for two-way interactions only.**
  Three-way and higher-order terms appear in the HR table and the
  effect-modification test, but are skipped for the subgroup breakdown.
- **Disabled in competing-risks (Fine-Gray) mode.** When Multiple Event
  Levels analysis is set to "Competing risks", within-subgroup refits are
  disabled (weighted subdistribution refits are fragile); the output notes
  this and directs you to the interaction coefficient instead.
- **Non-converged subgroups are flagged.** If a subgroup refit fails to
  converge (typically small-sample separation), its row is marked with an
  asterisk and a footnote - treat that HR with extreme caution.

## In R

The GUI builds exactly the Cox interaction model you would write directly
with the `survival` package:

```{r r-equivalent}
library(survival)

# A minimal simulated trial: does Treatment benefit differ by Biomarker status?
set.seed(1)
n <- 300
dat <- data.frame(
  Treatment = factor(sample(c("Control", "Drug"), n, replace = TRUE)),
  Biomarker = factor(sample(c("Negative", "Positive"), n, replace = TRUE))
)
lp <- with(dat, -0.1 * (Treatment == "Drug") +
                 -0.9 * (Treatment == "Drug" & Biomarker == "Positive"))
dat$time  <- rexp(n, rate = exp(lp) * 0.1)
dat$event <- rbinom(n, 1, 0.8)

# Treatment (focal) x Biomarker (moderator): the interaction test
fit <- coxph(Surv(time, event) ~ Treatment * Biomarker, data = dat)
summary(fit)

# Within-subgroup HR for Treatment, computed the same way multisurvival does:
# relevel Biomarker to each level and re-read the Treatment coefficient.
fit_neg <- coxph(Surv(time, event) ~ Treatment * relevel(Biomarker, ref = "Negative"), data = dat)
fit_pos <- coxph(Surv(time, event) ~ Treatment * relevel(Biomarker, ref = "Positive"), data = dat)
```

`multisurvival`'s Interaction (Effect-Modification) Test table reports the
same HR/CI/p as the `Treatment:BiomarkerPositive` row of `summary(fit)`
above, and its Within-Subgroup Hazard Ratios table reports the
`TreatmentDrug` coefficient from each releveled refit.
